Notes on AWI

### 2023-10-21 repro/papers/awi/paper.tex
update

AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
The extended modeling operator ${\bar F}$ maps extended sources $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure field. That is, the extended source depends on the receiver location as well as the source location - so there is one acoustic system for each source *and* receiver position - a lot of wave equations! If all of the wavelets for each source are the same, that is, ${\bar w}({\bf x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of $F$. This is the *source-recever* extension, in the terminology of Huang et al. 2015.

AWI assumes that the extended sources are time convolutions of the (known) exact source with an {\em adaptive kernel} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time. Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
$$
\bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
$$
For notational convenience, introduce the notation $S[m,w_*]$ for the convolution operator with kernel $F[m]w_*$:
\begin{equation}
  \label{eqn:defs}
  S[m,w_*]u = (F[m]w_*)*u
\end{equation}
Warner and Guasch assume (implicitly) that there is always a (near-)zero-residual solution of the extended inversion problem. That is, they assume that for any $m$, there is an extended source $\bar{w}$, or equivalently an adaptive kernel $u_0$ with $\bar{w} = w_**u_0$ for which $\bar{F}[m]\bar{w} = S[m,w_*]u_0\approx d$. If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$. Such a kernel $u_0$ is (approximately) in the null space of multiplication by $t$.

@@

\begin{itemize}
\item[1. ]Given $m$, solve the deconvolution problem $(F[m]w_*)*u_0 = d$ for $u_0$, if necessary in a least squares sense and with Tihonov regularization;. 
\item[3. ]Compute the objective 
  \begin{equation}
    \label{eqn:tJz}
    \tJz[m] = \sum_{x_s,x_r} \left(\frac{\int dt |tu_0|^2}{\int dt |u_0|^2}\right)
  \end{equation}
and its gradient, then update $\kappa,\beta$ using a gradient descent method to minimize $\tJz$.
\end{itemize}

The normalization of the integrand per source and receiver by the $L^2$ norm of $u_0$ is quite important, as will be explained shortly.

@@

For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$,, suppressed from the notation,
\begin{equation}
  \label{eqn:hadamard}
  F[m]w_*(t) = a[m]w_*(t-\tau[m]) + Rw_*(t)
\end{equation}
where $R(\bx_r,\bx_s)$ is a bounded operator on $L^2((-\infty,T])$ bounded by a multiple of $\lambda(w_*)$. This ``single arrival'' approximation  follows from the expansion of the fundamental solution of the wave equation due to Hadamard, see for instance \cite{Friedlander:75}. This construction is sketched in Appendix B. Note that in general the approximation \ref{eqn:hadamard} is only local: for smooth $m$, it holds for $\bx_r$ sufficiently close to $\bx_s$, but fails at larger distances.

For approximations with $O(\lambda(w_*)$ remainders, write
\begin{equation}
  \label{eqn:pwe}
  F[m]w_*(t) \approx a[m]w_*(t-\tau[m])
\end{equation}
Thanks to the approximation \ref{eqn:pwe},
\begin{equation}
  \label{eqn:spwe}
S[m,w_*]u(t) \approx a[m](w_**u)(t-\tau[m])
\end{equation}
Assume that the data is noise-free:
$$
d(t)=F[m_*]w_*(t) \approx a[m_*]w_*(t-\tau[m_*])
$$
The (approximate) solution of $S[m,w_*]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$. There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined. To fix this, it's necessary to treat $u_0$ as the solution of a regularized least squares problem: define
$$
u_0^{\epsilon} = \mbox{argmin}_u (\|S[m,w_*]u-d\|^2 + \epsilon^2\|u\|^2)
$$
$$
\approx \mbox{argmin}_u (\|a[m](w_**u)(\cdot-\tau[m])-a[m_*]w_*(\cdot-\tau[m_*])\|^2 + \epsilon^2\|u\|^2)
$$
Use the temporary abbreviations $a_*=a[m_*], a=a[m], \tau_*=\tau[m_*], \tau=\tau[m]$, and write in terms of the Fourier transforms $\hat{u}_{\epsilon},\hat{w_*}$ of $u_0^{\epsilon}$ and $w_*$:
$$
\hat{u}_{\epsilon} \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w_*}\hat{u}e^{i\omega \tau}-a_*\hat{w_*}e^{i\omega\tau_*}|^2 + \epsilon^2 |\hat{u}|^2\right)
$$
The normal equation is
$$
(a^2 |\hat{w}_*|^2 +\epsilon^2)\hat{u} = a a_*|\hat{w}_*|^2e^{i\omega(\tau_*-\tau)}
$$
the solution of which is
$$
\hat{u}_{\epsilon} \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\epsilon} e^{i\omega(\tau_*-\tau)}
$$
where
$$ 
\hat{g}_{\epsilon} = \frac{a_*^2|\hat{w}_*|^2}{a_*^2|\hat{w}_*|^2 + \epsilon^2}.
$$
Since $w_* \in C_0^{\infty}$, $\hat{w_*}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\epsilon}: \epsilon > 0$ tends to $1$ almost everywhere as $\epsilon \rightarrow 0$. Hence the inverse Fourier transform $g_{\epsilon}$ tends to $\delta$ in the sense of distributions, and 
$$
u_0^{\epsilon}(t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution solution of $\bar{S}[m,w_*]u=d$. However for $\epsilon>0$, $u_0^{\epsilon}$ is square integrable.

@@

\section{A Penalty Version of AWI}
The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem. This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
\begin{equation}
  \label{eqn:Ja}
  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m,w_*]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
\end{equation}
Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
\[
  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
\]
define
\[
  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
\]
and
\begin{equation}
  \label{eqn:defmnorm}
  \|u\|^2_m = \langle u, W u \rangle.
\end{equation}
With this notation,
\[
  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
\]
In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
\begin{equation}
  \label{eqn:tJa}
  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
\end{equation}
This is the so-called variable projection reduction of the quadratic $\Ja$.
Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.

@@

Apply this relation in the context of the AWI penalty problem defined earlier. $M$ is an open set in $C^k(\Omega, \bR^2)$. $U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,. $S$ is the regularized modeling operator $(S[m,w_*],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.

@@

>>>>

AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
The extended modeling operator ${\bar F}$ maps extended sources
$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
field. That is, the extended source depends on the receiver location
as well as the source location - so there is one acoustic system for
each source *and* receiver position - a lot of wave equations! If all
of the wavelets for each source are the same, that is, ${\bar w}({\bf
  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.

AWI assumes that the extended sources are time convolutions of the
(known) exact source $w_*$ with an {\em adaptive filter} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time. Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
\[
 \bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
\]
Since the known source $w_*$ is not adjusted during the inversion, 
introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w_*$:
\begin{equation}
  \label{eqn:sdef}
  \equiv S[m]u
\end{equation}
Warner and Guasch assume (implicitly) that there is always a
(near-)zero-residual solution of the extended inversion problem. That
is, they assume that for any $m$, there is an extended source
$\bar{w}$, or equivalently an adaptive filter $u_0$ with $\bar{w}=w_**u_0$, for which $S[m]u_0\approx d$. If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$. Such a kernel $u_0$ is (approximately) in the null space of multiplication by $t$.

@@

\begin{itemize}
\item[1. ]Given $m$, solve the deconvolution problem $S[m]u_0 = d$ for
  $u_0$, if necessary in a least squares sense and with Tihonov
  regularization (``prewhitening'');
\item[3. ]Compute the objective 
  \begin{equation}
    \label{eqn:tJz}
    \tJz[m] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_0(\bx_r,t;\bx_s)|^2}{\int dt |u_0(\bx_r,t;\bx_s)|^2}\right)
  \end{equation}
and its gradient, then update $\kappa,\beta$ using a gradient descent method to minimize $\tJz$.
\end{itemize}

The trace-by-trace normalization of the integrand by the
$L^2$ norm of $u_0$ is quite important, as will be explained shortly.

@@

For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$, suppressed from the notation,
\begin{equation}
  \label{eqn:hadamard}
  F[m]w_*(t) = a[m]w_*(t-\tau[m]) + Rw_*(t)
\end{equation}
where $R(\bx_r,\bx_s)$ is a bounded operator on $L^2((-\infty,T])$ bounded by a multiple of $\lambda(w_*)$. This ``single arrival'' approximation  follows from the expansion of the fundamental solution of the wave equation due to Hadamard, see for instance \cite{Friedlander:75}. This construction is sketched in Appendix B. Note that in general the approximation \ref{eqn:hadamard} is only local: for smooth $m$, it holds for $\bx_r$ sufficiently close to $\bx_s$, but fails at larger distances.

For approximations with $O(\lambda(w_*)$ remainders, write
\begin{equation}
  \label{eqn:pwe}
  F[m]w_*(t) \approx a[m]w_*(t-\tau[m])
\end{equation}
Thanks to the approximation \ref{eqn:pwe},
\begin{equation}
  \label{eqn:spwe}
S[m]u(t) \approx a[m](w_**u)(t-\tau[m])
\end{equation}
Assume that the data is noise-free:
$$
d(t)=F[m_*]w_*(t) \approx a[m_*]w_*(t-\tau[m_*])
$$
The (approximate) solution of $S[m]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$. There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined. To fix this, it's necessary to treat $u_0$ as the solution of a regularized least squares problem: define
$$
u_0^{\epsilon} = \mbox{argmin}_u (\|S[m]u-d\|^2 + \epsilon^2\|u\|^2)
$$
$$
\approx \mbox{argmin}_u (\|a[m](w_**u)(\cdot-\tau[m])-a[m_*]w_*(\cdot-\tau[m_*])\|^2 + \epsilon^2\|u\|^2)
$$
Use the temporary abbreviations $a_*=a[m_*], a=a[m], \tau_*=\tau[m_*], \tau=\tau[m]$, and write in terms of the Fourier transforms $\hat{u}_{\epsilon},\hat{w_*}$ of $u_0^{\epsilon}$ and $w_*$:
$$
\hat{u}_{\epsilon} \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w_*}\hat{u}e^{i\omega \tau}-a_*\hat{w_*}e^{i\omega\tau_*}|^2 + \epsilon^2 |\hat{u}|^2\right)
$$
The normal equation is
$$
(a^2 |\hat{w}_*|^2 +\epsilon^2)\hat{u} = a a_*|\hat{w}_*|^2e^{i\omega(\tau_*-\tau)}
$$
the solution of which is
$$
\hat{u}_{\epsilon} \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\epsilon} e^{i\omega(\tau_*-\tau)}
$$
where
$$ 
\hat{g}_{\epsilon} = \frac{a_*^2|\hat{w}_*|^2}{a_*^2|\hat{w}_*|^2 + \epsilon^2}.
$$
Since $w_* \in C_0^{\infty}$, $\hat{w_*}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\epsilon}: \epsilon > 0$ tends to $1$ almost everywhere as $\epsilon \rightarrow 0$. Hence the inverse Fourier transform $g_{\epsilon}$ tends to $\delta$ in the sense of distributions, and 
$$
u_0^{\epsilon}(t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution solution of $\bar{S}[m,w_*]u=d$. However for $\epsilon>0$, $u_0^{\epsilon}$ is square integrable.

@@

\section{A Penalty Version of AWI}
The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem. This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
\begin{equation}
  \label{eqn:Ja}
  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
\end{equation}
Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
\[
  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
\]
define
\[
  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
\]
and
\begin{equation}
  \label{eqn:defmnorm}
  \|u\|^2_m = \langle u, W u \rangle.
\end{equation}
With this notation,
\[
  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
\]
In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
\begin{equation}
  \label{eqn:tJa}
  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
\end{equation}
This is the so-called variable projection reduction of the quadratic $\Ja$.
Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.

@@

Apply this relation in the context of the AWI penalty problem defined earlier. $M$ is an open set in $C^k(\Omega, \bR^2)$. $U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,. $S$ is the regularized modeling operator $(S[m],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.

@@

\section{Computation of the Reduced Penalty Function and its Gradient}
Like all penalty functions, this one is best viewed as a least-squares objective of a map with values in a product space: define
$$
A_{\alpha,\epsilon}[m] =  
\left[
\begin{array}{c}
S[m]\\
\epsilon I\\
\alpha T_{\epsilon}[m]
\end{array}
\right]
, \, b = \left[
\begin{array}{c}
d\\
0\\
0
\end{array}
\right]
$$
Then
$$
J_{\alpha,\epsilon}[m,w_*,u] = 0.5\|A_{\alpha,\epsilon}[m]u-b\|^2
$$

Both Newton's method with the Kaufman Hessian approximation (described in the VCL notebook) and LBFGS or similar first-order methods apply to minimization of $\tilde{J}_{\alpha,\epsilon}[m]$. As shown in the VCL notebook, application of these algorithms to a VPM reduction like $\tilde{J}_{\alpha,\epsilon}$ requires access to two linear operator valued functions of $m$, namely $A_{\alpha,\epsilon}[m]$ and the (partial) derivative $D_m(A_{\alpha,\epsilon}[m]u)$. The former is the main actor in computation of the objective via solution of the normal equations. The latter appears in the VPM gradient formula:
$$
\nabla \tilde{J}_{\alpha,\epsilon}[m] = (D_m A_{\alpha,\epsilon}[m]u)|_{u=\tilde{u}_{\alpha,\epsilon}[m]}^T(A_{\alpha,\epsilon}[m]\tilde{u}_{\alpha,\epsilon}[m] -b)
$$
Note that $D_m A_{\alpha,\epsilon}[m]u$ is a map from the model parameter space to the product range space of $A$, and the transpose is a map in the other direction.

\subsection{Computing $D_mA_{\alpha,\epsilon}[m]^T$}
The main goal of this and the next two subsections is to express the
operator in the subsection title in terms of
\begin{itemize}
  \item trace-by-trace convolution and cross-correlation operators,
  \item the derivative $D_mF[m]w_*$ and its adjoint, and
  \item trace-by-trace scaling.
\end{itemize}
These the first two bullets describe common components of FWI.The
third is necessary to express the AWI penalty operator.

Note that the second-mentioned operator describes the derivative of
the acoustic forward modeling operator for a single point source per
receiver gather, not the extended modeling operator $\bar{F}$. The
latter is not a modular component of FWI, and a self-contained
implementation is not likely to be available.

Write $\nabla \tilde{J}_{\alpha,\epsilon}$ in terms of its components:
\begin{equation}
= D_mS[m]^T e_0 + \alpha D_mT_{\epsilon}[m]^Te_2
\label{eqn:tildejgrad}
\end{equation}
where $e=(e_0,e_1,e_2)^T$ is the residual vector:
$$
e = \left[
\begin{array}{c}
S[m]u-d\\
\epsilon u\\
\alpha T[m]u
\end{array}
\right],\, u=\tilde{u}_{\alpha,\epsilon}[m].
$$
Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.

Evidently the key calculations are $D_mS^T$ and $D_m T^T$.

\subsection{Convolution Revisited}

To keep track of the data flow in subsequent calculations, it is convenient to introduce a notation for the operator of convolution by $u$. This notation must include the time ranges for input and output traces, as these define the index limits in the sums that implement the convolution operator. Looked at another way, these ranges define the domain and range of the operator, and that is how the definition in *segyvc.ConvolutionOperator* is arranged: the arguments to the class constructor are the domain and range (as *segyvc.Spaces*, type-checked) and the filename for the convolution kernel. Domain, range, and kernel must share spatial geometry (source and receiver positions, in particular number of traces, and time step; these conditions are checked. In this section, the convolution operator with compatible trace space domain $X$ and range $Y$, and kernel $z \in Z$ is denoted $C[X,Y,Z,z]$. The kernel space $Z$ is also compatible with $X$ and $Y$.

The domain, range, and kernel spaces considered here are all coordinate spaces, defined by finite time index ranges (or, in the continuum case, bounded time ranges). For such a coordinate space $X$, denote by $\Pi_X$ the orthogonal projection of the ambient space $l^2$ (or $L^2({\bf R})$ onto $X$. In the discrete case, this projection $\Pi_X$ simply extracts the values inside the characterizing index range $I_X$:
$$
(\Pi_X f)_i = f_i, \, i \in I_X,
$$
and similarly for the continuum case. The transpose projection $\Pi_X^T$ injects the array of values within the range into $l^2$, with zeros outside. So $\Pi_X^T\Pi_X$ is the projection onto the subspace of $l^2$ (or $L^2({\bf R})$) corresponding to $X$, i.e. the range of $\Pi_X^T$, and $\Pi_X \Pi_X^T = I_X$.

Define convolution $f*g$ by the usual sum or integral, for arguments $f$ and $g$ of bounded support. With these conventions,
$$
C[X,Y,Z,z]x = \Pi_Y(\Pi_Z^T z * \Pi_X^T x).
$$
The class *segyvc.ConvolutionOperator* implements precisely the discrete version of this definition.

Both discrete and continuous convolution are commutative. The same is true of the discrete, finite version of convolution implemented in *segyvc.ConvolutionOperator*, and can be expressed as the follow obvious consequence of the definition: if $X, Y, Z$ are compatible trace spaces, and $z \in Z, x \in X$, then
\begin{equation}
C[X,Y,Z,z]x = C[Z,Y,X,x]z
\label{eqn:convcomm}
\end{equation}
The transpose of convolution is expressed via the time reversal operator $f \mapsto \check{f}\, \check{f}_i = f_{-i}$:  
$$
\langle f*g,h \rangle = \langle g, \check{f}*h \rangle.
$$ 
The right-hand side defines the cross-correlation of $f$ and $h$. From the definition of $C$,
$$
\langle C[X,Y,Z,z]x, y\rangle_Y = \langle \Pi_Y(\Pi_Z^T z * \Pi_X^T x), y \rangle_Y
$$
$$
= \langle \Pi_Z^T z * \Pi_X^T x, \Pi_y^T y \rangle = \langle \Pi_X^T x, \check{(\Pi_Z^T z)}* \Pi_Y^T y \rangle
$$
Denote by $\check{Z}$ the coordinate space produced by time reversal applied to $Z$, that is, $I_{\check{Z}}$ consists of the negatives of members of $I_Z$. Then $\check{(\Pi_Z^T z)}=\Pi_{\check{Z}}\check{z}$, so the above is
$$
=\langle \Pi_X^T x, \Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y \rangle =\langle x, \Pi_X(\Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y) \rangle_X
$$
$$
= \langle x, C[Y,X,\check{Z},\check{z}]y \rangle_X.
$$
That is, $C[X,Y,Z,z]^T = C[Y,X,\check{Z},\check{z}]$.

Combining the commutativity relation $\ref{eqn:convcomm}$ with this identity yields the useful identity:
\begin{equation}
C[X,Y,Z,z]^Ty = C[Y,X,\check{Z},\check{z}]y =
C[\check{Z},X,Y,y]\check{z} =
C[X,\check{Z},\check{Y},\check{y}]^T\check{z}.
\label{eqn:convcommtransp}
\end{equation}

\subsection{Back to AWI penalty gradient}

The SEGY trace spaces involved in the AWI calculation are the data space $D$ and the adaptive filter space $U$. Minimization of $J_{\alpha,\epsilon}$ amounts to solution of a least squares problem with operator $C[U,D,D,F[m]w]$, that is, convolution with the predicted data:
$$
S[m]u = C[U,D,D,F[m]w_*]u = C[D,D,U,u]F[m]w_*
$$
The second expression, equivalent by the commutativity relation \ref{eqn:convcomm} to the first, is more convenient for computing the derivative with respect to $m$ of $S[m]u$:
$$
D_m(S[m]u) \delta m = C[D,D,U,u](D_mF[m]\delta m)
$$
whence ($e_0 = S[m]u_{\alpha,\epsilon}[m]-d \in D$) the first summand in the gradient of $\tilde{J}_{\alpha,\epsilon}$ is
$$
D_m(S[m]u)^T e_0 = D_mF[m]^TC[D,D,U,u]^Te_0.
$$
%The derivative $D_mF[m]$ and its adjoint are defined in the *vcl.Function* subclass *asg.fsbop*, and the convolution operator $C$ and its transpose in *segyvc.ConvolutionOperator*.

Introduce the set $SR$ of source-receiver pairs active in the data
$D$, and scale factor space $P = \{f:SR \rightarrow \bR\}$. Write
\[
  T[m]u = B_2 \circ B_1 \circ B_0[m],
\]
where 
\[
B_2: P \times U \rightarrow U, \, B_2[f,u](bx_r,t;\bx_s) = f(\bx_r,\bx_s) tu(\bx_r,t;\bx_s)
\]
\[
B_1: U \rightarrow P, \,B_1[u](\bx_r,t,\bx_s) = \|u(\bx_r,\cdot; \bx_s)\|^{-1}
\]
\[
B_0: M \rightarrow U, \, B_0[m] = u_{0,\epsilon}[m]
\]
Then
\[
 \frac{\alpha^2}{2} D_m(\|T[m]u\|_U^2)\delta m = \alpha^2\langle D_m(
 T[m]u)\delta m, T[m]u \rangle_U
\]
\[
  =\alpha^2 \langle DB_2(B_1\circ B_0[m],u)DB_1(B_0[m]) DB_0[m]\delta m,
  T[m]u\rangle_U
\]
\begin{equation}
  \label{eqn:compgrad}
  =\alpha^2 \langle DB_0[m]\delta m, DB_1(B_0[m])^TDB_2(B_1\circ
  B_0[m],u)^TT[m]u\rangle_U
\end{equation}
Next display the derivative and adjoint derivative of $B_1$ and $B_2$:
$$
(D_fB_2[f,u]\delta f)(\bx_r,t;\bx_s) = tu(\bx_r,t;\bx_s)\delta f(\bx_r,\bx_s)
$$
$$
D_f(B_2[f,u])^Te_2(\bx_r,\bx_s) = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_S);
$$
$$
D_u B_1[u]\delta u(\bx_r,\bx_s) = -\|u(\bx_r,\cdot;\bx_s)\|^{-3}\langle u(bx_r,\cdot;\bx_s), \delta u(\bx_r,\cdot;\bx_s) \rangle
$$
$$
(D_u B_1[u])^Tf(\bx_r,t;\bx_s) =
-f(\bx_r,\bx_s)\|u(\bx_r,\cdot;\bx_s\|^{-3}u(\bx_r,t;\bx_s)
$$
With these expressions, compose various components appearing on the
right hand side of equation \ref{eqn:compgrad}:
\[
  B_1\circ B_0[m] (\bx_r,\bx_s) =  \|\tilde{u}_{0,\epsilon}(\bx_r,\cdot; \bx_s)\|^{-1},
\]
\[
 DB_2(B_1\circ B_0[m],u)^Te_2 = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s),
\]
(Note that the result is independent of the first argument.)
\[
DB_1(B_0[m])^Tf(\bx_r,t:\bx_s)=
-f(\bx_r,\bx_s)\|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s),
\]
and
\[
  DB_1(B_0[m])^TDB_2(B_1\circ  B_0[m],u)^TT[m]u =
\]
\begin{equation}
  \label{eqn:compgrad1}
  -\int dt\,t(T[m]u)(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s)
  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
\end{equation}
$$
u_{0,\epsilon}[m] = (S[m]^TS[m] + \epsilon^2 I)^{-1}S[m]^Td
$$
$$
= (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}C[U,D,D,F[m]w_*]^Td.
$$
So
$$
D_mB_0[m]\delta m = (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}
$$
$$
\times ((D_mC[U,D,D,F[m]w_*]\delta m)^Td -(D_m(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*])\delta m)u_{0,\epsilon}[m])
$$
$$
=(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}
$$
$$
\times
(D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m  - C[U,D,D,F[m]w_*]^TD_m(C[U,D,D,F[m]w_*]u_0)\delta m)
$$ 
where the derivatives are to be evaluated before the insertion of the definitions 
$$
r_0 = d -C[U,D,D,F[m]w_*]u_{0,\epsilon}[m], \, u_0 = u_{0,\epsilon}[m]
$$
From the commutativity relation,
$$
D_m(C[U,D,D,F[m]w_*]u_0)\delta m = D_m(C[D,D,U,u_0]F[m]w_*)\delta m
=C[D,D,U,u_0]D_mF[m]\delta m.
$$
Similarly, use the identity \ref{eqn:convcommtransp} to write
$$
C[U,D,D,F[m]w_*]^Tr_0 = C[U,\check{D},\check{D},\check{r_0}]^T({F[m]w_*})\check, 
$$
so
$$
D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m=
C[U,\check{D},\check{D},\check{r_0}]^T(D_mF[m]w_* \delta m)\check.
$$

Putting these calculations together,
